Data Specialization for a Staged Lambda Calculus

### 2015-02-28 paper/intro.tex
more terminology shifts
Editing pass on abstract (to tighten a bit)
tightened up 2.2

Programs featuring this structure are particularly important in real-time
computer graphics; in fact, modern graphics architectures \emph{require} that
graphics computations be organized into distinct {\em passes} which perform
increasingly fine-grained computations (for example, per object, per region of
the screen, and per pixel, where many computations in later passes use the
results from a single computation in an earlier pass)\,\cite{OpenGL4Spec}.

While performant, this paradigm of programming with explicit passes results in
complex code where invariants must hold across different passes, and local
changes for one pass may require changes to other passes to ``plumb'' inputs
from all other frequencies\,\cite{Foley:2011}.
%
In other words, requiring manual separation of passes breaks key programming
abstractions such as composition and modularity.
%
Graphics researchers therefore have suggested using explicitly staged programs
\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring to the compiler the
mechanical task of separating such programs into passes.

@@

There has been much work on language-based techniques for performing
precomputation. Partial evaluation~\cite{futamura71,jones96} achieves
precomputation by specializing functions to specific arguments. Going back to
our example, if we specialize $f$ to a particular $v$, written $f_v$, such that
$f(v,y) = f_v(y)$, then we can specialize $F$ to $v$ as
\begin{lstlisting}
fun @$F_v(y_1, y_2, \ldots, y_m)$@ = 
  @$f_v(y_1);  f_v(y_2); \ldots; f_v(y_m)$@.
\end{lstlisting}
eliminating the need to compute $m$ times those parts of $f(v,-)$ which do not
depend on the second argument.

@@

We have implemented this splitting algorithm (\ref{sec:implementation}) and
applied it to a number of staged programs (\ref{sec:examples}) ranging from
straight-line arithmetic operations to recursive and higher-order functions. In
the case of \texttt{quickselect}, which we discuss next, the split code is in
fact asymptotically faster over many $y_i$.

>>>>

Programs featuring this structure are particularly important in real-time
computer graphics; in fact, modern graphics architectures \emph{require} that
graphics computations be organized into distinct {\em passes} which perform
increasingly fine-grained computations (for example, per object, per region of
the screen, and per pixel, where many computations in later passes use the
results from a single computation in an earlier pass)\,\cite{OpenGL4Spec}.
While performant, this paradigm of programming with explicit passes results in
complex code where invariants must hold across different passes, and local
changes for one pass may require changes to other passes to ``plumb'' inputs
from all other frequencies\,\cite{Foley:2011}.
%
In other words, requiring manual separation of passes breaks key programming
abstractions such as composition and modularity.
%
Graphics researchers therefore have suggested using explicitly staged programs
\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring to the compiler the
mechanical task of separating such programs into passes.

@@

There has been much work on language-based techniques for performing
precomputation. Partial evaluation~\cite{futamura71,jones96} achieves
precomputation by specializing functions to specific arguments. Going back to
our example, if $f$ specializes to a particular $v$, written $f_v$, such that
$f(v,y) = f_v(y)$, then $F$ can be specialized to $v$ as
\begin{lstlisting}
fun @$F_v(y_1, y_2, \ldots, y_m)$@ = 
  @$f_v(y_1);  f_v(y_2); \ldots; f_v(y_m)$@.
\end{lstlisting}
eliminating the need to compute $m$ times those parts of $f(v,-)$ which do not
depend on the second argument.

@@

We have implemented this splitting algorithm (\ref{sec:implementation}) and
applied it to a number of staged programs (\ref{sec:examples}) ranging from
straight-line arithmetic operations to recursive and higher-order functions. In
the case of the quickselect algorithm, which we discuss next, the split code is
in fact asymptotically faster over many $y_i$.

### 2015-02-28 paper/overview.tex
tightened up 2.2

A more sophisticated solution is to, in a first pass, construct a binary search
tree out of \texttt{l}, and in a second pass, simply look for the \texttt{k}th
leftmost element of that tree. To do this efficiently requires one more
innovation---storing at each node the size of its left subtree (i.e., using
\emph{augmented trees}). This approach has (expected) runtime $\Theta(n\log{n} +
m\log{n})$, a nearly linear-time improvement in $m$.

@@

In this paper, we present a splitting algorithm which transforms staged programs
into multi-pass programs. In certain cases, this algorithm \emph{automatically}
improves algorithms in the way described above.

@@

\texttt{qss} is obtained from \texttt{qs} by wrapping certain computations with
\texttt{prev} and \texttt{next}, signaling transitions between first- and
second-stage code, and $\texttt{g}$, signaling ground terms.  
%
We also use a function \texttt{hold},
\lstinline{1`hold : ^int -> $`2`int`},
to promote first-stage integers to second-stage integers.
%
Our type system ensures that the staging annotations in \texttt{qss} are
consistent, in the sense that computations marked as first-stage cannot depend
on ones marked as second-stage.

@@

The staged quickselect code shown in \ref{fig:qs-staged} makes
explicit the staging of all terms, making it natural to ask, whether
we can transform this code into manually implemented multi-pass code
described above and match the efficiency improvements of the manual
implementation. Considering the algorithmic knowledge and the
considerations needed, this may seem like a tall order.
Interestingly, our splitting algorithm achieves exactly this.
In the rest of this section, we present a brief, high-level overview
of the main ideas behind this algorithm using the quickselect example.
%

When applied to the staged code in \ref{fig:qs-staged}, our splitting
algorithm yields a two-pass program that uses the binary-search-tree
based implementation.  Specifically, in its first pass, the two-pass
program takes the input list and constructs a probabilistically
balanced binary search tree, which is isomorphic to a treap data
structure~\cite{SeidelAr96}.  In the second pass, the program performs,
for each rank, a binary search tree lookup, by walking the tree to
find the element with the desired rank.
%
\ref{fig:qs-staged} illustrates the code for the first and second
passes \texttt{qs1} and \texttt{qs2} of the staged function
\texttt{qss}.

To create the multi-pass algorithm, the splitting algorithm operates
by composing local transformations on the subterms of the input
program.  In particular, the algorithm has no special knowledge of the
quickselect algorithm, binary search trees, or how to perform lookups
on binary search trees, but it is able to derive them from the input
program.
%

The splitting algorithm scans the program code for first-stage
computations (which depend only on first-stage values) and separates
them into the function, \texttt{qs1} of the first pass. This function
performs the first-stage computations and places the results into a
boundary data structure that both records the control flow and the
results from the first stage at relevant control-flow points.  More
specifically, the function \texttt{qs1} performs all the recursive
calls and evaluates all instances of the partition function, which
depend only on the input list.  The function produces a boundary data
structure that collects the results from all recursive calls along
with a tag that indicates the control branches taken.  Since
\texttt{qss} has a binary control structure (casing on the list), the
boundary is a binary tree.  To ensure that the trace contains the
necessary results to complete the execution in the second pass, where
second-stage values may be used, the splitting algorithm includes in
the nodes of the tree information such as the "pivot" used for
splitting and the size of the subtree at that note.  The resulting
tree is thus not just a tree but a binary search tree, keyed by the
pivot and augmented with "size" information.

As the splitting algorithm scans \texttt{qss} for computations that
can be performed in the first stage, it also collects computations
that must be left to the second stage in a separate function
\texttt{qs2}.  This function, which is executed in the second pass,
takes as argument the boundary and the second-stage argument (the
rank) and performs a lookup in the boundary data structure.  But how
does \texttt{qs2} knows to traverse the tree? It does not. Since
the splitting algorithm recorded the control flow of the first-stage
execution in the boundary data structure, \texttt{qs2} simply follows
this control structure and performs at each point the parts of the
computation from \texttt{qss} that can now be performed in the
presence of the second-stage argument (the rank).  In the context of
our example, this performs a lookup on the boundary data structure by
using the supplied rank.

\begin{theorem}
  Consider an execution of \texttt{qs1} with a randomly permuted input
  of $n$ keys and performing $m$ executions of \texttt{qs2} with the
  result of \texttt{qs1}.  The total run-time for this computation is
  $\Theta(n\log{n} + m\log{n})$ in expectation.
\end{theorem}
\begin{proof}
  We use a few standard facts.  We note first that \texttt{qs1} is
  isomorphic to a quick-sort algorithm but builds a binary search tree
  instead of combining the returned results in a list.  The run-time
  for \texttt{qs1} is thus (expected) $\Theta(n\log{n})$.
%
  The depth of the tree generated by \texttt{qs1} is isomorphic to the
    depth of a run of the quickselect algorithm or equivalently to the
    span of a parallel implementation of the quicksort algorithm, both
    of which are known to be (expected) $\Theta(\log{n})$.
%
    By inspection of \texttt{qs2}, we can see that the algorithm takes
    time proportional to the depth of the binary search tree, which is
    $\Theta(\log{n})$. Consider now a random variable for each of the
    $m$ invocations of \texttt{qs2}, each of which takes expected
    $\Theta(\log{n})$ time.  Since the total time is the sum of these
    random variables and since the expectation of the sum is sum of
    the expectations of the random variables, the bound follows.
\end{proof}

>>>>

A more sophisticated solution is to, in a first pass, construct a binary search
tree out of \texttt{l}, and in a second pass, simply look for the \texttt{k}th
leftmost element of that tree. To do this efficiently requires one more
innovation---augmenting the tree by storing at each node the size of its left
subtree. This approach has (expected) runtime $\Theta(n\log{n} + m\log{n})$, a
nearly linear-time improvement in $m$.

@@

In this paper, we present a splitting algorithm which automatically transforms
staged programs into multi-pass programs. In certain cases, this algorithm even
improves input algorithms in the way described above!

@@

\texttt{qss} is obtained from \texttt{qs} by wrapping certain computations with
\texttt{prev} and \texttt{next}, signaling transitions between first- and
second-stage code, and $\texttt{gr}$, signaling ground terms.  
%
We also use a function
%\lstinline{1`hold : ^int -> $`2`int`},
to promote first-stage integers to second-stage integers.
%
Our type system ensures that the staging annotations in \texttt{qss} are
consistent, in the sense that computations marked as first-stage cannot depend
on ones marked as second-stage.

@@

In the rest of this section, we present a high-level overview of the main ideas
behind our splitting algorithm, applied to \texttt{qss}.
%
Splitting \texttt{qss} yields a two-pass program that creates a
probabilistically balanced, augmented binary search tree (which is isomorphic to
a treap data structure~\cite{SeidelAr96}) as an intermediate data structure. In
particular, its first pass (\texttt{qs1}) constructs such a binary search tree,
and its second pass (\texttt{qs2}) traverses the tree, using the embedded size
information to find the element of the desired rank.
%
The code for \texttt{qs1} and \texttt{qs2} is in \ref{fig:qs-staged}.

%Our splitting algorithm is guided by the local structure of the input program.
%In particular, the algorithm has no special knowledge of the quickselect
%algorithm, binary search trees, or how to perform lookups on binary search
%trees, but it is able to derive them from the input program.

Our splitting algorithm scans \texttt{qss} for first-stage computations,
gathering them into \texttt{qs1}. Given \texttt{l}, this pass performs these
computations and places the information needed by the subsequent pass into a
boundary data structure.
%
In particular, \texttt{qs1} performs all recursive calls and evaluates all
instances of \texttt{partition} (since it depends only on \texttt{l}). It
produces a boundary data structure that collects the results from these
recursive calls, tagged by the branch (\texttt{LT}, \texttt{EQ}, or \texttt{GT})
in which that call occurred. Since the recursive calls occur in two different
branches (\texttt{LT} and \texttt{GT}) the boundary structure is a binary tree.  
%
Lastly, it records \texttt{i} (the size of the left subtree) and \texttt{\#1 ht}
(the pivot/head of the list) in the boundary structure, because those
computations are \emph{held} for use in the second stage.
%
The final result is a binary search tree augmented with size information, and
whose keys are the pivots.

Our splitting algorithm simultaneously scans \texttt{qss} for second-stage
computations, gathering them into \texttt{qs2}. This pass is given the boundary
data structure and the rank \texttt{k}, and finishes the computation. Now that
\texttt{k} is known, the conditional on \texttt{compare k i} can be evaluated,
choosing which recursive call of \texttt{qss} is actually relevant for this
\texttt{k}. Since the boundary data structure contains the first-stage data for
\emph{all} of the recursive calls, performing these comparisons essentially
walks the tree, using the rank along with the size data \texttt{i} in order to
look up the \texttt{k}th leftmost node in the tree.

\begin{theorem}
  Consider an execution of \texttt{qs1} with a randomly permuted input
  of $n$ keys and performing $m$ executions of \texttt{qs2} with the
  result of \texttt{qs1}.  The total run-time for this computation is
  $\Theta(n\log{n} + m\log{n})$ in expectation.
\end{theorem}
\begin{proof}
  We use a few standard facts.  We note first that \texttt{qs1} is
  isomorphic to a quicksort algorithm but builds a binary search tree
  instead of combining the returned results in a list.  The run-time
  for \texttt{qs1} is thus (expected) $\Theta(n\log{n})$.
%
  The depth of the tree generated by \texttt{qs1} is isomorphic to the
    depth of a run of the quickselect algorithm, or equivalently to the
    span of a parallel implementation of the quicksort algorithm, both
    of which are known to be (expected) $\Theta(\log{n})$.
%
    By inspection of \texttt{qs2}, we can see that the algorithm takes
    time proportional to the depth of the binary search tree, which is
    $\Theta(\log{n})$. Consider now a random variable for each of the
    $m$ invocations of \texttt{qs2}, each of which takes expected
    $\Theta(\log{n})$ time.  Since the total time is the sum of these
    random variables, and the expectation of the sum is the sum of
    the expectations of the random variables, the bound follows.
\end{proof}
